Morse Theory

### 2020-11-29 paper.tex
merge

\textbf{Orientable Manifolds} \\
Our goal is to express number of loops in a Reeb graph by # of critical points. Let $n_i$ be the number of nodes with degree $i$. Orientable manifolds have 0 $n_2$ because every point is either minimum, maximum or saddle.\\
Therefore the number of arcs is number of connections each point has, divided by two (because we count each arc twice) is $e = (n_1+3n_3)/2$ \\ 
number of loops is $1+e-(n_1+n_3)$
\begin{definition}
    Let P be a polyhedron with V vertices, E edges, and F faces. Then we define the Euler characteristic to be 
    \begin{equation}
        \chi(P) = V - E + F  
    \end{equation}
\end{definition} \\
More generally, Euler characteristic is defined as alternating sum of Betti numbers. Last Morse inequality allow us to define $\chi$ as a set of inequalities for alternating sums of Betti numbers in terms of a corresponding alternating sum of the number of critical point $n_i$ of a Morse function of a given index.
\begin{lemma}
    The Reeb graph of a Morse function on a connected, orientable 2-manifold of genus g has g loops.
    \begin{proof}
        We first suppose that Reeb graph has no loops. It has $n_1 = n_3+2$. And if we write $c_i$ for the number of critical points of index $i$ we have $n_i = c_0+c_2$ and $n_3 = c_1$. we have $\chi = c_1 - c_1 +c_2 = n_1 - n_3 = 2.$  2 is the Euler characteristic of a sphere, which has no genus. \\ 
        Now suppose that Reeb graph has at least one loop. We can apply following two functions because they preserve homotopy type and the number of loops: collapse degree 1 nodes and merge arcs across degree 2 nodes which get eliminated in the process. For example consider the following graph:
    \begin{center}
    \includegraphics[width=0.4\textwidth]{homotopic_transformation.png}
    \end{center}
    Let $m_3$ be the number of remaining degree 3 nodes and note that it is even because $3m_3$ is twice the number of remaining arcs. Using the Euler-Poincare Theorem for graphs we get $\chi = m_3 - 3m_3/2 = \beta_0 - \beta_1$. The graph is connected which implies that the number of loops is $\beta = m_3/2+1$. We have $c_1$ degree 3 nodes in the original Reeb graph and for each minimum and maximum we collapse one degree 1 node removing a 3 degree node in the process. Last strong Morse inequality tells us that $m_3 = c_1 - (c_0 + c_2) = -\chi = 2g - 2$. The number of loops is therefore $ \beta_1 = (2g-2)/2+1=g$
    \end{proof}
\end{lemma}
% % end M2
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% % begin V2
% % end V2
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{thebibliography}{9}
     \bibitem{s}
         {\sc }
         `` https://core.ac.uk/download/pdf/20670562.pdf
\end{thebibliography}

\caption{ 4 figures}

\textbf{Non-orientable 2-manifolds without boundary}\\

@@

\end{document}

>>>>

\subsection{Orientable Manifolds} 
Our goal is to express number of loops in a Reeb graph by  of critical points. Let $n_i$ be the number of nodes with degree $i$. Orientable manifolds have 0 $n_2$ because every point is either minimum, maximum or saddle.\\
Therefore the number of arcs is number of connections each point has, divided by two (because we count each arc twice) is $e = (n_1+3n_3)/2$ \\ 
number of loops is $1+e-(n_1+n_3)$
\begin{definition}
    Let P be a polyhedron with V vertices, E edges, and F faces. Then we define the Euler characteristic to be 
    \begin{equation}
        \chi(P) = V - E + F  
    \end{equation}
\end{definition} 
More generally, Euler characteristic is defined as alternating sum of Betti numbers. Last Morse inequality allow us to define $\chi$ as a set of inequalities for alternating sums of Betti numbers in terms of a corresponding alternating sum of the number of critical point $n_i$ of a Morse function of a given index.
\begin{lemma}
    The Reeb graph of a Morse function on a connected, orientable 2-manifold of genus g has g loops.
    \begin{proof}
        We first suppose that Reeb graph has no loops. It has $n_1 = n_3+2$. And if we write $c_i$ for the number of critical points of index $i$ we have $n_i = c_0+c_2$ and $n_3 = c_1$. we have $\chi = c_1 - c_1 +c_2 = n_1 - n_3 = 2.$  2 is the Euler characteristic of a sphere, which has no genus. \\ 
        Now suppose that Reeb graph has at least one loop. We can apply following two functions because they preserve homotopy type and the number of loops: collapse degree 1 nodes and merge arcs across degree 2 nodes which get eliminated in the process. For example consider the following graph:
    \begin{center}
    \includegraphics[width=0.4\textwidth]{homotopic_transformation.png}
    \end{center}
    Let $m_3$ be the number of remaining degree 3 nodes and note that it is even because $3m_3$ is twice the number of remaining arcs. Using the Euler-Poincare Theorem for graphs we get $\chi = m_3 - 3m_3/2 = \beta_0 - \beta_1$. The graph is connected which implies that the number of loops is $\beta = m_3/2+1$. We have $c_1$ degree 3 nodes in the original Reeb graph and for each minimum and maximum we collapse one degree 1 node removing a 3 degree node in the process. Last strong Morse inequality tells us that $m_3 = c_1 - (c_0 + c_2) = -\chi = 2g - 2$. The number of loops is therefore $ \beta_1 = (2g-2)/2+1=g$
    \end{proof}
\end{lemma}
% % end M2
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% % begin V2
% % end V2
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \caption{ 4 figures}

\subsection{Non-orientable 2-manifolds without boundary}

@@

\begin{thebibliography}{9}
     \bibitem{s}
         {\sc }
         `` https://core.ac.uk/download/pdf/20670562.pdf
\end{thebibliography}
\end{document}
